---
title: "MA415 Midterm Project"
author: "Kristen Bestavros"
date: "2025-03-27"
format:
  pdf: default
execute:
  echo: false
editor: visual
---

## Midterm Project: Analyzing Strawberries in Florida and California

This project explores chemical usage and strawberry production patterns in California and Florida using USDA NASS data. Our focus is twofold: first, to compare conventional chemical usage practices across states, and second, to investigate yield, price, and revenue differences for organic strawberries. The analysis incorporates survey and census data from 2021 and 2023.

We start with data cleaning and preparation, which involves reading the data from the CSV, splitting census and survey data, aggregating by state, and a few other data processing steps that will make things easier later. Once data cleaning is complete, we will start the assignment by comparing chemical usage on strawberries in Florida and California.

## Data Prep

```{r}
#| label: load libraries
#| warning: false
#| message: false


library(knitr)  
library(kableExtra)
library(tidyverse)
library(stringr)
library(dplyr)
library(ggrepel)
library(scales)
library(GGally)
```

```{r}
#| label: read data - glimpse
#| warning: false


strawberry <- read_csv("strawb_mar6.csv", 
                       col_names = TRUE,
                       show_col_types = FALSE)

source("my_functions.R")

strawb <- strawberry |> drop_one_value_col()

show_unique <- function(data, nrows=10 ){
  # make a tibble items to hold the data to show
  # browser()
  a <- nrows * dim(data)[2]  # number of cells in items
  items <- rep(" ", a) # items will coerce everything to char
  dim(items) <- c(nrows ,dim(data)[2]) # shape items
  items <- as_tibble(items)
  colnames(items) <- colnames(data)
  # browser()
  for(i in 1:dim(data)[2]){

    col_items <- unique(data[,i])
    # row_ex is the number of rows needed 
    # to make the column length conformable with items
    row_ex <- nrows - dim(col_items)[1] 
    if(row_ex >= 0){
      ex_rows <- tibble(rep(" ",row_ex))
      colnames(ex_rows) <- colnames(col_items)
      col_add <- rbind2(col_items, ex_rows)
      
    } else if(row_ex < 0){
      col_add <- col_items[1:10,]
      
    }

    items[,i] <- col_add

  }
  
  return(items)
}
```

```{r}
#|label: split strawb into census and survey pieces
#|warning: false

strw_census <- strawb |> filter(Program == "CENSUS")

strw_survey <- strawb |> filter(Program == "SURVEY")

nrow(strawb) == (nrow(strw_census) + nrow(strw_survey))

```

```{r}
#| label: examine Census and survey tibbles
#| warning: false


s_census <- strw_census |> drop_one_value_col(prt_val = TRUE)

s_survey <- strw_survey |> drop_one_value_col(prt_val = TRUE)


unique_sur <- s_survey |> show_unique(nrows = 10)

unique_cen <- s_census |> show_unique(nrows = 10)


strw_census <- s_census |> select(-`State ANSI`)

strw_survey <- s_survey |> select(-`State ANSI`, -`Week Ending`, -Period)

rm(s_census, s_survey, strawberry, strawb, items)


```

```{r}
#| label: work on Census columns
#| warning: false

# Save a copy of the original for safety
strw_census_orig <- strw_census

# Check column names
names(strw_census)

# First split Commodity
strw_census <- strw_census |>
  separate_wider_delim(
    cols = Commodity,
    delim = ",",
    names = c("INCOME", "NET_CASH_FARM", "STRAW"),
    too_few = "align_start",
    too_many = "drop"
  )

strw_census <- strw_census |>
  separate_wider_delim(
    cols = `Data Item`,
    delim = ",",
    names = c("Fruit", "Category", "Item"),
    too_few = "align_start",
    too_many = "merge"
  )


```

```{r}
#| label: Assignment shortner results
#| warning: false

straw_cen_f <- strw_census |> filter(State == "FLORIDA")

straw_sur_f <- strw_survey |> filter(State == "FLORIDA")
straw_cen_c <- strw_census |> filter(State == "CALIFORNIA")
straw_sur_c <- strw_survey |> filter(State == "CALIFORNIA")

rm(strw_census, strw_survey, unique_cen, unique_sur)

```

## **Methodology and Decision Rationale**

**Chemical Selection:** I chose my three chemicals based on coverage across both states and agricultural relevance:

-   **Abamectin**, an insecticide/miticide

-   **Cyprodinil** and **Fludioxonil**, which are frequently co-applied (e.g., in Switch®)

These choices allowed us to contrast insect vs. fungal control and observe whether California's higher production scale translated into greater chemical reliance.

**Graph Type Choices:**

-   **Bar Charts** visualize raw chemical usage by year and state

-   **Stacked Bars** summarize aggregate chemical inputs by category

-   **Proportion Charts** remove magnitude bias to compare relative usage

-   **Percent Change Bars** capture trends across time

-   **Scatter Plot** highlights the correlation between Cyprodinil and Fludioxonil

-   **Lollipop Charts** and **summary tables** communicate pricing and revenue relationships clearly

All graphs are faceted or colored by state to allow easy visual comparison.

## Chemical Usage Comparison: Abamectin, Cyprodinil, and Fludioxonil

To explore chemical usage patterns in strawberry production, we focus on three commonly applied pesticides: **Abamectin**, **Cyprodinil**, and **Fludioxonil**. These were selected based on their appearance in both California and Florida datasets, their agricultural importance, and their narrative potential.

Together, this trio allows us to contrast **insect control versus fungal control**, **cross-state usage patterns**, and even **chemical pairings**.

```{r}
#| label: Choose and contrast unique chemicals
#| warning: false

# Combine all four into one data frame
straw_all <- bind_rows(
  straw_sur_c,
  straw_sur_f,
  straw_cen_c,
  straw_cen_f
)

target_chems <- c(
  "CHEMICAL, FUNGICIDE: (CYPRODINIL = 288202)",
  "CHEMICAL, FUNGICIDE: (FLUDIOXONIL = 71503)",
  "CHEMICAL, INSECTICIDE: (ABAMECTIN = 122804)"
)

chem_data <- straw_all |>
  filter(`Domain Category` %in% target_chems) |>
  mutate(Suppressed = Value == "(D)",
         Value = ifelse(str_detect(Value, "^[0-9,]+$"),
                        as.numeric(gsub(",", "", Value)),
                        NA),
         Chemical = case_when(
            str_detect(`Domain Category`, "CYPRODINIL") ~ "Cyprodinil",
            str_detect(`Domain Category`, "FLUDIOXONIL") ~ "Fludioxonil",
            str_detect(`Domain Category`, "ABAMECTIN") ~ "Abamectin"
    ))

chem_data_clean <- chem_data |> drop_na(Value)

ggplot(chem_data, aes(x = factor(Year), y = Value, fill = State)) +
  geom_col(position = "dodge") +
  facet_wrap(~ Chemical, scales = "free_y") +
  labs(title = "Chemical Usage by State and Year",
       x = "Year", y = "Usage (units in Value column)") +
  theme_minimal()
```

**Figure 1: Chemical Usage by State and Year** California uses substantially more of all three chemicals, especially fungicides. This likely reflects larger-scale production and greater post-harvest preservation needs. Florida increased its use of Abamectin in 2023, possibly indicating rising pest pressure.

The large gap between states in **Cyprodinil** and **Fludioxonil** usage could be due to differences in disease prevalence, shipping requirements (fungicides help reduce post-harvest decay), or product availability.

```{r}
#| label: proportional chemical comparison
#| warning: false


# Only keep rows with numeric Value and both states represented
prop_data <- chem_data |> 
  drop_na(Value) |> 
  group_by(Chemical, Year) |> 
  filter(n_distinct(State) == 2) |>  # Only keep years with both states
  mutate(Share = Value / sum(Value))

# Plot
ggplot(prop_data, aes(x = factor(Year), y = Share, fill = State)) +
  geom_col(position = "fill") +
  facet_wrap(~ Chemical) +
  scale_y_continuous(labels = scales::percent_format()) +
  labs(title = "Proportion of Chemical Usage by State",
       y = "Share of Total Usage", x = "Year") +
  theme_minimal()


```

**Figure 2: Proportional Usage by State** To remove the influence of scale and highlight relative usage, we visualize the proportional share of each chemical used by state. For Cyprodinil and Fludioxonil, California consistently accounts for 60–75% of total reported usage in both years. Interestingly, the percentage of Cyprodinil or Fludioxonil used by Florida decreased between 2021 and 2023, yet Florida's usage of Abamectin as an insecticide increased drastically. There may be a policy or environmental reason for this – we will look at it again in the conclusion.

```{r}
#| label: stacked chemical usage
#| warning: false

# Filter and clean chemical usage data
chem_stacked <- chem_data_clean %>%
  group_by(State, Year, Chemical) %>%
  summarise(Total = sum(Value), .groups = "drop")

# Plot: Stacked bars by chemical for California and Florida
ggplot(chem_stacked, aes(x = factor(Year), y = Total, fill = Chemical)) +
  geom_col() +
  facet_wrap(~State) +
  labs(
    title = "Stacked Chemical Usage by State and Year",
    x = "Year",
    y = "Total Usage",
    fill = "Chemical"
  ) +
  theme_minimal()

```

**Figure 3: Stacked Usage by State** Total fungicide application surged in California between 2021 and 2023, driven by Cyprodinil. Florida's increase was more moderate, but the composition shifted slightly toward fungicides.

```{r}
#| label: percent change chemical summary
#| warning: false

# Summarize total usage by Chemical, State, and Year
chem_summary <- chem_data_clean |>
  group_by(Chemical, State, Year) |>
  summarise(Total_Usage = sum(Value), .groups = "drop")

# Pivot to wide format: separate 2021 and 2023 into columns
chem_wide <- chem_summary |>
  filter(Year %in% c(2021, 2023)) |>  # Only focus on years with data
  pivot_wider(names_from = Year, values_from = Total_Usage, 
              names_prefix = "Year_") |>
  drop_na()

# Calculate absolute and percent change
chem_change <- chem_wide |>
  mutate(
    Absolute_Change = Year_2023 - Year_2021,
    Percent_Change = 100 * (Year_2023 - Year_2021) / Year_2021
  )

# Plot percent change (from 2021 to 2023)
library(ggplot2)

ggplot(chem_change, aes(x = Chemical, y = Percent_Change, fill = State)) +
  geom_col(position = "dodge") +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray40") +
  geom_text(
    aes(label = paste0(round(Percent_Change, 1), "%")),
    position = position_dodge(width = 0.9),
    vjust = ifelse(chem_change$Percent_Change >= 0, -0.5, 1.5),
    size = 3.5
  ) +
  scale_y_continuous(labels = scales::percent_format(scale = 1)) +
  labs(
    title = "Change in Chemical Usage from 2021 to 2023",
    x = "Chemical",
    y = "Percent Change from 2021",
    fill = "State"
  ) +
  theme_minimal()
```

**Figure 4: Percent Change in Chemical Usage** Florida's Abamectin usage exploded (+1528%), starting from a small base, as we previously observed in figure 2. California saw steady increases in all three, notably a +197% jump in Fludioxonil.

```{r}
#| label: chemical usage table
#| warning: false


chem_change |>
  select(Chemical, State, Year_2021, Year_2023, Absolute_Change, Percent_Change) |>
  mutate(across(c(Year_2021, Year_2023, Absolute_Change), round, 0),
         Percent_Change = round(Percent_Change, 1)) |>
  arrange(Chemical, State) |>
  kable(
    caption = "Chemical Usage in 2021 and 2023 with Absolute and Percent Change",
    col.names = c("Chemical", "State", "2021 Usage", "2023 Usage", "Change", "Percent Change"),
    align = c("l", "r", "r", "r", "r", "r")
  ) |>
  kable_styling(full_width = FALSE, bootstrap_options = c("striped", "hover"))
```

**Figure 5: Chemical Usage Table** The table confirms trends: California's values are orders of magnitude higher. Still, Florida’s absolute increases are notable given its smaller industry size.

```{r}
#| label: Cyprodinil and fludioxonil correlation
#| warning: false


cypro_fludiox <- chem_data |> 
  filter(Chemical %in% c("Cyprodinil", "Fludioxonil")) |>
  drop_na(Value) |>
  group_by(Year, State, Chemical) |>
  summarise(Total = sum(Value), .groups = "drop") |>
  pivot_wider(names_from = Chemical, values_from = Total) |> 
  drop_na()

ggplot(cypro_fludiox, aes(x = Cyprodinil, y = Fludioxonil, color = State)) +
  geom_point(size = 3) +
  geom_smooth(method = "lm", se = FALSE) +
  labs(title = "Correlation Between Cyprodinil and Fludioxonil Usage",
       x = "Cyprodinil (units)", y = "Fludioxonil (units)") +
  theme_minimal()


```

**Figure 6: Correlation Between Cyprodinil and Fludioxonil** A strong linear relationship supports the hypothesis that these fungicides are co-applied. This applies in both states, though California uses much more overall.

## **Strawberry Price and Volume Comparison: Focus on Organic**

In the second half of the analysis, we focus on organic strawberries to explore production efficiency and market trends. Organic data were more consistently reported and include clear sales and acreage information, making it possible to derive additional metrics. Specifically, we calculated:

-   **Yield (CWT per Acre)** = Volume / Acres

-   **Revenue per Acre** = Revenue / Acres

-   **Price per CWT** = Revenue / Volume

These derived metrics allow us to compare not just raw outputs, but production efficiency and profitability across California and Florida. While both states grow organic strawberries, their practices and market strategies differ significantly. Organic strawberries also offer a cleaner lens for comparison, since they exclude some of the chemical usage confounding present in conventional data.

```{r}
#| label: Organic data prep
#| warning: false

organic_data <- straw_all %>%
  filter(str_detect(Category, "ORGANIC"))

organic_data <- organic_data %>%
  mutate(
    Category = str_trim(Category),
    Value_num = as.numeric(gsub(",", "", Value))
  ) %>%
  filter(str_detect(Category, "ORGANIC")) %>%
  mutate(
    Metric = case_when(
      str_detect(Category, "ACRES HARVESTED") ~ "Acres",
      str_detect(Category, "SALES") & str_detect(Item, "\\$") ~ "Revenue",
      str_detect(Category, "SALES") & str_detect(Item, "CWT") ~ "Volume",
      TRUE ~ NA_character_
    )
  ) %>%
  filter(!is.na(Metric)) %>%
  select(Year, State, Metric, Value_num) %>%
  pivot_wider(names_from = Metric, values_from = Value_num)

organic_metrics <- organic_data %>%
  mutate(
    Yield_CWT_per_Acre = Volume / Acres,
    Revenue_per_Acre = Revenue / Acres,
    Price_per_CWT = Revenue / Volume
  )
```

```{r}
#| label: Organic strawberry price lollipop graph
#| warning: false


lollipop_data <- organic_metrics %>%
  pivot_longer(cols = c(Revenue_per_Acre, Price_per_CWT), names_to = "Metric", values_to = "Value")

ggplot(lollipop_data, aes(x = State, y = Value, color = State)) +
  geom_segment(aes(xend = State, y = 0, yend = Value), size = 1) +
  geom_point(size = 4) +
  facet_wrap(~Metric, scales = "free_y") +
  labs(
    title = "Organic Strawberry Price and Revenue by State",
    y = "Value",
    x = ""
  ) +
  theme_minimal() +
  theme(legend.position = "none")
```

**Figure 7: Organic Price and Revenue by State** Florida commands a higher price per CWT, but California achieves significantly more revenue per acre due to its much higher yields.

```{r}
#| label: Organic yield and price per state
#| warning: false


ratio_data <- organic_metrics %>%
  select(State, Yield_CWT_per_Acre, Price_per_CWT) %>%
  pivot_longer(cols = c(Yield_CWT_per_Acre, Price_per_CWT), names_to = "Metric", values_to = "Value")

ggplot(ratio_data, aes(x = Metric, y = Value, fill = State)) +
  geom_col(position = "dodge") +
  labs(
    title = "Yield and Price Comparison per State (Organic Strawberries)",
    x = "",
    y = "Value"
  ) +
  theme_minimal()
```

**Figure 8: Organic Yield and Price Comparison** The bar chart clearly shows Florida’s price advantage but also its yield disadvantage. California’s yield per acre is more than triple Florida’s.

```{r}
#| label: Organic summary table
#| warning: false


organic_summary <- organic_metrics |>
  group_by(State, Year) |>
  summarise(
    Avg_Price_CWT = mean(Price_per_CWT, na.rm = TRUE),
    Avg_Yield = mean(Yield_CWT_per_Acre, na.rm = TRUE),
    Avg_Revenue = mean(Revenue_per_Acre, na.rm = TRUE),
    .groups = "drop"
  )

# Display as a table
organic_summary |>
  mutate(across(starts_with("Avg_"), round, 1)) |>
  kable(
    caption = "Average Organic Strawberry Metrics by State and Year",
    col.names = c("State", "Year", "Avg. Price per CWT", "Avg. Yield (CWT/Acre)", "Avg. Revenue per Acre"),
    align = c("l", "c", "r", "r", "r")
  ) |>
  kable_styling(full_width = FALSE, bootstrap_options = c("striped", "hover"))

```

**Figure 9: Summary Table - Organic Metrics** Florida's 2021 average price per CWT is \$273.40 compared to California’s \$220.70, but yield and revenue lag far behind. California earns \$73,742 per acre versus Florida’s \$26,077.

## **Conclusion and Insights**

California's strawberry industry is not just larger but also characterized by intensive chemical usage, particularly fungicides. This aligns with its role as the top strawberry-producing state in the U.S., responsible for nearly 90% of all strawberries grown domestically (USDA, 2022). The scale of California's production likely necessitates greater pest and disease management, particularly to meet strict export and shelf-life standards (Bolda et al., 2017). This explains the higher levels of Cyprodinil and Fludioxonil used in California compared to Florida.

Florida's smaller-scale strawberry industry, while still significant, appears to focus on earlier-season market windows and fresh-market premiums. Its notably higher price per CWT for organic strawberries may reflect these marketing advantages, as well as a shorter supply chain and regional demand. However, Florida's significantly lower yields limit its per-acre revenue. This suggests that while Florida benefits from higher unit prices, it lacks the production scale and efficiency of California.

The explosive rise in Abamectin usage in Florida may indicate changing pest pressures or evolving integrated pest management strategies, possibly in response to warmer winters and increased mite activity (Smith & Noling, 2021). Meanwhile, California’s continued investment in fungicide co-application likely reflects systemic practices built into commercial operations, packing houses, and export logistics.

Organic strawberry metrics further underscore the economic advantage of California’s high-yield systems. Despite earning less per unit, California’s total revenue per acre far surpasses Florida’s, a result of substantially higher productivity.

Taken together, these patterns suggest that climate, market orientation, infrastructure, and agronomic conditions all influence how states manage chemical inputs and achieve profitability in strawberry production.

*References:*

-   USDA National Agricultural Statistics Service (2022). "Strawberries, Annual Summary."

-   Bolda, M., Klonsky, K., & De Moura, R. (2017). "Sample Costs to Produce and Harvest Strawberries: Central Coast. UC Cooperative Extension."

-   Smith, H. & Noling, J. (2021). "Strawberry Production in Florida." University of Florida IFAS Extension.

### **Limitations and Considerations**

-   **Missing Data:** Many cells were suppressed ("(D)") or reported inconsistently across years.

-   **Metric Derivation:** Yield and revenue were inferred from other fields, introducing compounded uncertainty.

-   **Survey vs. Census Blending:** Merging these sources may hide reporting biases.

-   **No Acre-Specific Chemical Data:** Without knowing how many acres were treated, usage figures may exaggerate or understate per-acre intensity.

Despite these challenges, the patterns revealed are consistent, logical, and well-supported by the data available.

*All data from USDA NASS. Charts and analysis generated in R using tidyverse and ggplot2. Coding support and discussion guidance provided by ChatGPT.*
